---
title: "Reproducible_Research_CourseProject2"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

  Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.


## Synopsis

  The goal of the assignment is to explore the NOAA Storm Database and explore the effects of severe weather events on both population and economy.The database covers the time period between 1950 and November 2011.

## Data Processing

Download the raw data file and extract the data 

```{r}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl, destfile = paste0("F:/datasciencecoursera/Repro_Project2", '/repdata%2Fdata%2FStormData.csv.bz2'))
```

Loading the data and extract it into a dataframe and Subset the storm data according to damage on Health and Economy

```{r}
Storm_df <- read.csv("F:/datasciencecoursera/Repro_Project2/repdata%2Fdata%2FStormData.csv.bz2")

S_health_data <-subset(Storm_df, select=c(EVTYPE,FATALITIES,INJURIES))
S_economy_data <-subset(Storm_df, select=c(EVTYPE,CROPDMG,CROPDMGEXP,PROPDMG,PROPDMGEXP))
```

Loading ggplot2,ggpubr library

```{r}
library(ggplot2)
library(ggpubr)
```

## Data Cleaning

Checking the factorlevels of the PROPDMGEXP and CROPDMGEXP variables

```{r}
levels(factor(S_economy_data[,5]))
levels(factor(S_economy_data[,3]))
```

Cleaning the property damaged data

```{r}
S_economy_data$PROPDMGEXP[S_economy_data$PROPDMGEXP == ""] <- NA
S_economy_data$PROPDMGEXP <- gsub("1", "10", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("NA", "1", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("\\-", "1", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("\\?", "1", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("\\+", "1", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("^0$", "1", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("h|H", "100", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("K", "1000", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("m|M", "1000000", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("B", "1000000000", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("2", "100", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("3", "1000", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("4", "10000", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("5", "100000", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("6", "1000000", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("7", "10000000", S_economy_data$PROPDMGEXP)
S_economy_data$PROPDMGEXP <- gsub("8", "100000000", S_economy_data$PROPDMGEXP)
```

 Cleaning the crop damaged data
 
```{r}
S_economy_data$CROPDMGEXP[S_economy_data$CROPDMGEXP == ""] <- NA
S_economy_data$CROPDMGEXP <- gsub("NA", "1", S_economy_data$CROPDMGEXP)
S_economy_data$CROPDMGEXP <- gsub("\\?", "1", S_economy_data$CROPDMGEXP)
S_economy_data$CROPDMGEXP <- gsub("^0$", "1", S_economy_data$CROPDMGEXP)
S_economy_data$CROPDMGEXP <- gsub("k|K", "1000", S_economy_data$CROPDMGEXP)
S_economy_data$CROPDMGEXP <- gsub("m|M", "1000000", S_economy_data$CROPDMGEXP)
S_economy_data$CROPDMGEXP <- gsub("B", "1000000000", S_economy_data$CROPDMGEXP)
S_economy_data$CROPDMGEXP <- gsub("2", "100", S_economy_data$CROPDMGEXP)
```


## Results

### The impact on event types in Economy

##### Checking property damage by event types
 
Creating a dataset to summarize the property damage by event types.
Listing the top 10 with highest values.

```{r}
S_economy_data$TOTALPROP <- S_economy_data$PROPDMG * as.numeric(S_economy_data$PROPDMGEXP)
Storm_property <- aggregate(TOTALPROP ~ EVTYPE, data = S_economy_data, sum)
max_property <- Storm_property[order(-Storm_property$TOTALPROP), ]
max_property[1:10, ]
```

Visualizing the Top Ten of event types with the highest Property damage

```{r}
ggplot(data = max_property[1:10, ],aes(x=TOTALPROP,y=EVTYPE))+
  geom_bar(stat="identity" ,fill = "#34A287") +
  ggpubr::rotate_x_text() +
  coord_flip() + xlab("Total number of Property damage") +  ylab("Event Types") +
  ggtitle("Top Ten of event types with the highest Property damage")
```

##### Observation

The event type that produces the highest property damage in the U.S. is Flood
    
##### Checking crop damage by event types
    
Creating a dataset to summarize the crop damage by event types.
Ordering the results and selecting the top ten with the highest values

```{r}
S_economy_data$TOTALCROP <- S_economy_data$CROPDMG * as.numeric(S_economy_data$CROPDMGEXP)
Storm_crop <- aggregate(TOTALCROP ~ EVTYPE, data = S_economy_data, sum)
max_crop <- Storm_crop[order(-Storm_crop$TOTALCROP), ]
max_crop[1:10, ]
```

Visualizing the Top Ten of event types with the highest Crop damage

```{r}
ggplot(data = max_crop[1:10, ],aes(x=TOTALCROP,y=EVTYPE) )+
  geom_bar(stat="identity" ,fill = "#895178") +
  ggpubr::rotate_x_text() +
  coord_flip() + xlab("Total number of Crop damage") +  ylab("Event Types") +
  ggtitle("Top Ten of event types with the highest Crop damage")
```

##### Observation

  The event type that produces the highest crop damage in the U.S. is Drought

### The effect of event types in Population Health

Creating a dataset to summarize the fatalities by event type. 
Select top 10 having the highest values from the ordered result.

```{r}
Storm_fatalities <- aggregate(FATALITIES ~ EVTYPE, data = S_health_data, sum)
max_fatalities <- Storm_fatalities[order(-Storm_fatalities$FATALITIES), ]
max_fatalities[1:10, ]
```

Creating a dataset to summarize the injuries by event type.
Select top 10 having the highest values from the ordered result.

```{r}
Storm_injuries <- aggregate(INJURIES ~ EVTYPE, data = Storm_df, sum)
max_injuries <- Storm_injuries[order(-Storm_injuries$INJURIES), ] 
max_injuries[1:10, ]
```


Creating a dataset to summarize the fatalities + injuries by event type. 
Select top 10 having the highest values from the ordered result.

```{r}
S_health_data$HARM <- S_health_data$FATALITIES + S_health_data$INJURIES
Storm_harmful_total <- aggregate(HARM ~ EVTYPE, data = S_health_data, sum)
max_harmful <- Storm_harmful_total[order(-Storm_harmful_total$HARM), ]
max_harmful[1:10, ]
```

Visualizing the Top Ten of event types with the highest impact on Population Health

```{r}
ggplot(data = max_harmful[1:10, ],aes(x=HARM,y=EVTYPE) )+
  geom_bar(stat="identity" ,fill = "#A02A32") +
  ggpubr::rotate_x_text() +
  coord_flip() +  xlab("Total number of Fatalities + Injuries") + ylab("Event Types") +
  ggtitle("Top Ten of event types with the highest impact on Population Health")
```

##### Observation

  The most harmful event type to population health in the U.S. is Tornado
    
## Conclusions


  - The event type that produces the highest property damage in the U.S. is Flood.

  - The event type that produces the highest crop damage in the U.S. is Drought.

  - The most harmful event type to population health in the U.S. is Tornado.
  
  


